Measurement of a dwarf galaxy rotational velocity from its Lyman-alpha line emission

### 2016-09-23 paper/letter.tex
updated numbers and eqs

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Lyman-$\alpha$ Line Sketches a Supermassive Dwarf Galaxy}
\\
\\
Jaime E. Forero-Romero$^{1}$, Max Groenke$^2$, Mar\'ia Camila
Remolina-Guti\'errez$^1$, Juan Nicol\'as
Garavito-Carmargo$^3$, Mark Dijkstra$^2$.
\\
\\
\scriptsize
{$^1$ Departamento de Física, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia 
\\
$^2$ Institute of Theoretical Astrophysics, University of Oslo,
Postboks 1029 Blindern, NO-0315 Oslo, Norway.
\\
$^3$ Department of Astronomy, University of Arizona, 933 North Cherry
Avenue, Tucson, AZ 85721, USA. 
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ line emission that had evaded theoretical
interpretation so far. 
In this letter we explain these features by two different models: an
homogeneous gaseous sphere undergoing bulk rotation and an interstellar
medium composed by outflowing clumps with additional random motions.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these physical conditions.
We find that both models independently require high velocities
(either a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$ or a clump velocity
dispersion of \sigmaclump with outflows of
\inftyclump) consistent with a dynamical mass of at
least a billion solar masses, $7$ times larger than its baryonic mass.   
We argue that a possible explanation for this excess of
dynamical mass is the presence of a supermassive black hole at the
center of \tol. 
This work demonstrates the importance of considering multiphase
physics and rotation among the possible conditions shaping the
\lya\ spectra of the first galaxies.  
Additionally, if future kinematic maps of \tol\ confirm the high
velocities postulated in our model, it would provide new
evidence for dwarf galaxies as hosts of supermassive black
holes.  
}

@@

Assuming that the clumps are located in a spherical region of radius
$r_s=2.0$ kpc (corresponding to \tol's estimated 3D half-light
radius), 
this corresponds to dynamical masses of $M_{\rm dyn} =
2.8_{-0.8}^{+1.5}\times 10^{10} M_{\odot}$ and $M_{\rm
  dyn}=1.71\pm0.04 \times 10^{9}$ $M_{\odot}$ for the
rotation and multiphase models, respectively,

\tol's stellar mass is  $M_{\star} = 1.45\pm0.45\times 10^{8}
M_{\odot}$   \cite{2014PASP..126.1079M} and its total neutral HI mass is $M_{\rm HI}<2.65\times 10^{8}$ M$_{\odot}$
\cite{pustilnikmartin07}; the dynamical mass is at
least 7 to 68 times the baryonic mass, depending if one
considers the multiphase or rotation estimate.

We lean towards the lower dynamical mass estimate from the multiphase
model as it seems easier to reconcile with the following two
astrophysical mechanisms for its origin.
The first way to explain a dynamical mass of $10^{9} M_{\odot}$
in a sphere of 2 kpc could be having a dark matter halo of at
least $10^{12}$ M$_{\odot}$ in mass \cite{2011ApJ...726..108T}, which
leaves open the question as to why  \tol\ is not more similar to the
Milky Way galaxy as it is hosted by a dark matter halo of similar
mass. 
A second possibility is that \tol\ hosts a supermassive black hole of
$10^{9} M_{\odot}$. This is almost two orders of magnitude higher
than the supermassive black hole found in the compact dwarf galaxy
M60-UCD1 \cite{2014Natur.513..398S}, which has a similar stellar mass
as \tol. This also leaves open the question for the formation process
of such an system.

@@

We computed the projected half-luminosity radius to be $R_s=1.5\pm0.1$ kpc 
from the surface intensity profiles reported by.
\cite{2003A&A...410..481N}. 
Assuming spherical geometry, one can translate this value into a 3D
half-luminosity radius of $r_s=3R_s/2=2.0$ kpc.

@@

Both the rotation and the multiphase model constrain the typical
velocity $v$ of the HI gas, with and additional constrain on the typical
size for the emission region $r$ on could estimate a dynamical mass with

\begin{equation}
M_{\rm dyn} = \frac{v^{2}r}{G} = 1.16\times10^{9}
\left(\frac{v}{100\ \mathrm{km\ s}^{-1}}\right)^2\left(\frac{r}{\mathrm{kpc}}\right) M_{\odot}
\end{equation}

@@

>>>>

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Lyman-$\alpha$ Line Sketches a Supermassive Dwarf Galaxy}
\\
\\
Jaime E. Forero-Romero$^{1}$, Max Groenke$^2$, Mar\'ia Camila
Remolina-Guti\'errez$^1$, Juan Nicol\'as
Garavito-Carmargo$^3$, Mark Dijkstra$^2$.
\\
\\
\scriptsize
{$^1$ Departamento de Física, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia 
\\
$^2$ Institute of Theoretical Astrophysics, University of Oslo,
Postboks 1029 Blindern, NO-0315 Oslo, Norway.
\\
$^3$ Department of Astronomy, University of Arizona, 933 North Cherry
Avenue, Tucson, AZ 85721, USA. 
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ line emission that had evaded theoretical
interpretation so far. 
In this letter we explain these features by two different models: an
homogeneous gaseous sphere undergoing bulk rotation and an interstellar
medium composed by outflowing clumps with additional random motions.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these physical conditions.
We find that both models independently require high velocities
(either a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$ or a clump velocity
dispersion of \sigmaclump with outflows of
\inftyclump) consistent with a dynamical mass of at
least a billion solar masses, $6$ times larger than its baryonic mass.   
We argue that a possible explanation for this excess of
dynamical mass is the presence of a supermassive black hole at the
center of \tol. 
This work demonstrates the importance of considering multiphase
physics and rotation among the possible conditions shaping the
\lya\ spectra of the first galaxies.  
Additionally, if future kinematic maps of \tol\ confirm the high
velocities postulated in our model, it would provide new
evidence for dwarf galaxies as hosts of supermassive black
holes.  
}

@@

Assuming that the clumps are located in a spherical region of radius
$r_s=2.25$ kpc (corresponding to \tol's estimated 3D half-light
radius), 
this corresponds to dynamical masses of $M_{\rm dyn} =
3.2_{-1.0}^{+1.6}\times 10^{10} M_{\odot}$ and $M_{\rm
  dyn}=2.31\pm0.04 \times 10^{9}$ $M_{\odot}$ for the
rotation and multiphase models, respectively,

\tol's stellar mass is  $M_{\star} = 1.45\pm0.45\times 10^{8}
M_{\odot}$   \cite{2014PASP..126.1079M} and its total neutral HI mass is $M_{\rm HI}<2.65\times 10^{8}$ M$_{\odot}$
\cite{pustilnikmartin07}; the dynamical mass is at least 6 to 80 times
the baryonic mass, depending if one considers the multiphase or
rotation estimate.

We lean towards the lower dynamical mass estimate from the multiphase
model as it seems easier to reconcile with the following two
astrophysical mechanisms for its origin.
The first way to explain a dynamical mass of $10^{9} M_{\odot}$
in a sphere of 2 kpc could be having a dark matter halo of at
least $10^{12}$ M$_{\odot}$ in mass \cite{2011ApJ...726..108T}, which
leaves open the question as to why  \tol\ is not more similar to the
Milky Way galaxy as it is hosted by a dark matter halo of similar
mass. 
A second possibility is that \tol\ hosts a supermassive black hole of
$10^{9} M_{\odot}$. This is almost two orders of magnitude higher
than the supermassive black hole found in the compact dwarf galaxy
M60-UCD1 \cite{2014Natur.513..398S}, which has a similar stellar mass
as \tol. This would leave open the question about the formation
process of such a system.

@@

We computed the projected half-luminosity radius to be $R_s=1.5\pm0.1$ kpc 
from the surface intensity profiles reported by.
\cite{2003A&A...410..481N}. 
Assuming spherical geometry, one can translate this value into a 3D
half-luminosity radius of $r_s=3R_s/2=2.25$ kpc.

@@

Having constrains for  velocity dipersion $\sigma$ of some dynamical
tracers (clumps in the case of the multiphase model) in a spherical
system located in a region of size $r$ we estimate the dynamical mass
within $r$.

\begin{equation}
M_{\rm dyn} = 3 \frac{\sigma ^{2}r}{G} = 3.48\times10^{9}
\left(\frac{\sigma}{100\ \mathrm{km\ s}^{-1}}\right)^2\left(\frac{r}{\mathrm{kpc}}\right) M_{\odot}
\end{equation}

@@

In the case of rotational velocity $v$ in a region of size $r$ we
estimate the dynamical mass by

\begin{equation}
M_{\rm dyn} = \frac{v ^{2}r}{G} = 1.16\times10^{9}
\left(\frac{v}{100\ \mathrm{km\ s}^{-1}}\right)^2\left(\frac{r}{\mathrm{kpc}}\right)M_{\odot} 
\end{equation}
